A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-03-30 manuscript/manuscript.qmd

[**Competing Interests Statement:**]{.underline} Dr. Weberpals is now an employee of AstraZeneca and owns stocks in AstraZeneca. Dr. Kehl has received research funding from Meta, Inc. to his institution. Drs. Espirito and Robert are employees of McKesson and own McKesson stock. Dr. Wang has consulted ad hoc for Exponent Inc. and MITRE a federally funded research center for the Centers for Medicare and Medicaid Services on unrelated work. Dr. Schneeweiss is participating in investigator-initiated grants to the Brigham and Women's Hospital from Bayer and UCB unrelated to the topic of this study. He consults for and owns equity in Aetion Inc., a software manufacturer. He is an advisor to Temedica GmbH, a patient-oriented data generation company. His interests were declared, reviewed, and approved by the Brigham and Women's Hospital in accordance with their institutional compliance policies.

@@

Real-world evidence (RWE) is increasingly used to complement evidence from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice. Advancements in the curation and accessibility of electronic health record data (EHR) have presented the opportunity to utilize real-world data (RWD) to investigate therapeutic areas such as oncology, where administrative healthcare claims databases alone are often not fit-for-purpose. The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials in non-oncology indications. Here, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology. ENCORE is designed to emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. It will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to initiation and analysis. Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria. Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence the validity of oncological RWE studies.

@@

Randomized controlled trials (RCTs) are the go-to methodology for establishing the efficacy and safety of medical products. Under the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor consideration of real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 30% of new drug approvals, oncology was the disease area with the most FDA drug approvals in 2024[@senior2024fresh] and has many areas of high unmet medical need to treat serious conditions. Therefore, RWE has particularly important potential to complement evidence coming from RCTs in the field of oncology. Potential use cases include the assessment of effectiveness in specific patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials that have feasibility or equipoise challenges, or the precision oncology-focused discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between RWE studies and RCTs depends on multiple factors. Frequently referenced limitations include, lack of baseline randomization, missing data, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect. In settings where the RCT designs could be emulated well, RWE studies were able to reach the same conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.

@@

The rationale and operationalization of each criterion is listed in detail in @tbl-criteria. We will mainly consider pivotal large late-phase interventional RCTs after 2011 because treatment guidelines among included cancer indications have undergone significant changes in recent years. Due to the rapid adoption of breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in current clinical practice. Conversely, we excluded trials with results published too recently. To allow for enough data and follow-up time accrual in the real-world databases used for this project. Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not is often not reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges including missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest. Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.

>>>>

[**Competing Interests Statement:**]{.underline} Dr. Weberpals is now an employee of AstraZeneca and owns stocks in AstraZeneca. Dr. Kehl has received research funding from Meta, Inc. to his institution. Drs. Espirito and Robert are employees of McKesson and own McKesson stock. Dr. Wang has consulted ad hoc for Exponent Inc. and MITRE a federally funded research center for the Centers for Medicare and Medicaid Services on unrelated work. Dr. Glynn has received support for investigator-initiated grants to the Brigham and Women's Hospital from Amarin, AstraZeneca, Kowa, Novartis, and Pfizer unrelated to the current work. Dr. Schneeweiss is participating in investigator-initiated grants to the Brigham and Women's Hospital from Bayer and UCB unrelated to the topic of this study. He consults for and owns equity in Aetion Inc., a software manufacturer. He is an advisor to Temedica GmbH, a patient-oriented data generation company. His interests were declared, reviewed, and approved by the Brigham and Women's Hospital in accordance with their institutional compliance policies.

@@

Real-world evidence (RWE) is increasingly used to complement findings from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice. Advancements in the curation and accessibility of electronic health record data (EHR) present the opportunity to utilize real-world data (RWD) to investigate therapeutic areas including oncology, where administrative healthcare claims databases alone are often not fit-for-purpose. The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials in non-oncology indications. Here, we present the design and trial selection for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which extends this work to oncology. ENCORE is designed to emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. It will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to initiation and analysis. Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria. Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence the validity of oncological RWE studies.

@@

Randomized controlled trials (RCTs) are the go-to methodology for establishing the efficacy and safety of medical products. Directed by the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor consideration of real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 30% of new drug approvals, oncology was the disease area with the most FDA drug approvals in 2024[@senior2024fresh] and has many areas of high unmet medical need to treat serious conditions. Therefore, RWE has particularly important potential to complement evidence from RCTs in the field of oncology. Potential use cases include the assessment of effectiveness in specific patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials that have feasibility or equipoise challenges, or the precision oncology-focused discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between RWE studies and RCTs depends on multiple factors. Frequently referenced limitations include lack of baseline randomization, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect. In settings where the RCT designs could be emulated well, RWE studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.

@@

The rationale and operationalization for each criterion is listed in detail in @tbl-criteria. We will mainly consider pivotal large late-phase interventional RCTs after 2011 because treatment guidelines among included cancer indications have undergone significant changes in recent years. Due to the rapid adoption of breakthrough therapies in routine care, few patients currently initiate outdated treatment regimens in current clinical practice. Conversely, we excluded trials with results published too recently to allow for enough data and follow-up time accrual in the real-world databases used for this project. We did not define a global cut-off as the requirements for follow-up time are different for each cancer type and population (e.g., advanced NSCLC versus early-stage breast cancer) and the decisions were made on a case-by-case basis.

Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which is not well reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges including missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest. Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.
